---
title: "ISMB2021 Tutorial 8 - Workflow 2"
output: html_notebook
author: "Martina Summer-Kutmon"
date: "20 July 2021"
version: 1.0
comment: "adapted from Alexander Pico's Pathway Analysis Vignette"
license: "MIT License"
---

# General instructions (read before running the code snippets)
In this second workflow, we will create a protein-protein interaction network of the up- and down-regulated genes after COVID19 infection. Afterwards, we will extend the network with gene-pathway associations to see in which pathways the differentially expressed genes are present in.

* The script contains xxx code snippets which should be run one after the other. 
* Make sure all the required packages are installed beforehand (BiocManager::install(...)). 
* Make sure you have Cytoscape installed (version 3.8.0+) and running before you start running the script. 

*** 

# Setup

Loading libraries
```{r}
options(connectionObserver = NULL)

library(dplyr)
library(rWikiPathways)
library(RCy3)
library(RColorBrewer)
library(rstudioapi)
library(readr)
```

```{r}
setwd(dirname(getActiveDocumentContext()$path))
```

*** 

## Load differential gene expression dataset

Make sure you ran workflow 1 beforehand, so the differential gene expression file has been generated. 

We take the series 1 from the following dataset (NHBE mock treated versus SARS-CoV-2 infected): 
* https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE147507

which is related to the following publication:

Blanco-Melo, Daniel, et al. "Imbalanced host response to SARS-CoV-2 drives development of COVID-19." Cell 181.5 (2020): 1036-1045.

```{r}
data.NHBE <- read.table("data/deg-data.txt", sep = "\t", header=TRUE)
deg <- unique(data.NHBE[!is.na(data.NHBE$pvalue) & data.NHBE$pvalue < 0.05 & abs(data.NHBE$log2FoldChange) > 0.58,c(1,2)])
```

***

## PPI network analysis

Next, we will create a protein-protein interaction network with all differentially expressed genes using the STRING database. 

```{r}
RCy3::cytoscapePing()
installApp('stringApp') 

query <- format_csv(as.data.frame(deg$GeneId), col_names=F, quote_escape = "double", eol =",")
commandsRun(paste0('string protein query cutoff=0.9 newNetName="PPI network" query="',query,'" limit=0'))

# network will be opened in Cytoscape (this might take a while)
```

> Let's explore the network

- **Q1**: How many of the differentially expressed genes were found in STRING?
- **Q2**: Are all genes connected in the network? 
- **Q3**: Change the confidence cutoff in the commandsRun call from 0.9 (high confidence) to 0.4 (medium confidence). What changes? 

***

## Data visualization

Use the same visualization you created in workflow 1 to visualize the gene expression data on the network.

```{r}
loadTableData(data.NHBE, data.key.column = "GeneId", table.key.column = "query term")

# apply visual style with ?setNodeColorMapping
# create a gradient visualization for column: log2FoldChange
# Important add: style.name="STRING style v1.5"
# TODO!

setNodeColorMapping(...?...)
```

> Interpretation

- **Q4**: Do you see clusters of up- or down-regulated genes in the PPI network?

***

## Pathway information

Next, we will add information about participation of the differentially expressed genes in COVID19 specific pathways (Disease Map and COVID19 portal on WikiPathways).

```{r}
RCy3::cytoscapePing()
installApp('CyTargetLinker') 

pathways <- file.path(getwd(), "data/covid-pathways.xgmml")

# run CyTargetLinker
commandsRun(paste0('cytargetlinker extend idAttribute="query term" linkSetFiles="', pathways, '"'))
commandsRun('cytargetlinker applyLayout network="current"')
commandsRun('cytargetlinker applyVisualstyle network="current"')
RCy3::setNodeLabelMapping("display name", style.name="CyTargetLinker")

# there is an issue in the latest version with visualization of the added edges - the workaround below solves this for now
RCy3::cloneNetwork()
RCy3::setVisualStyle("default")
RCy3::setVisualStyle("CyTargetLinker")

# TODO: VISUAL STYLE
# adapt the visual style to also show the differential gene expression as the node fill color

```

> Interpretation

- **Q5**: How many differentially expressed genes are in at least one of the COVID pathways? 
- **Q6**: Are the genes also functionally related based on the PPI network?

***

## Save Cytoscape output and session

```{r}
# Saving output
png.file <- file.path(getwd(), "ppi-network.png")
exportImage(png.file,'PNG', zoom = 500)
cys.file <- file.path(getwd(), "ppi-network.cys")
saveSession(cys.file) 

#comment following line if you want to manipulate the visualization in Cytoscape
RCy3::closeSession(save.before.closing = F)
```